---
title: "project"
author: "David Hyon"
date: "2025-04-26"
output: pdf_document
---

# Load packages

```{r}
library(tigris)
library(tidyverse)
library(sf)
library(tidycensus)
library(lubridate)
library(rlang)
```




# Gameplan

# Election

## Read in 2024

```{r}
electionRaw <- read_csv("./2024_US_County_Level_Presidential_Results.csv")


election24 <- electionRaw %>% 
  select(county_fips, 1:9) %>% 
  mutate(
    trump = case_when(per_gop > per_dem ~ 1,
                      per_dem > per_gop ~ 0,
                      per_gop == per_dem ~ -999,
                      .default = NA),
    bell = if_else(trump == 1, 1, 0)
  )

table(election24$trump, useNA = c("always"))
rm(electionRaw)
```

## Read in 2020

```{r}
electionRaw <- read_csv("./2020_US_County_Level_Presidential_Results.csv")


election20 <- electionRaw %>% 
  # select(county_fips, 1:9) %>% # Keeps state and county name
  select(county_fips, 4:9) %>%  # Only voting data
  mutate(
    trump = case_when(per_gop > per_dem ~ 1,
                      per_dem > per_gop ~ 0,
                      per_gop == per_dem ~ -999,
                      .default = NA),
    bell = if_else(trump == 0, 1, 0)
  )

table(election20$trump, useNA = c("always"))
rm(electionRaw)
str(election20)
```
## Read in 2016

```{r}
electionRaw <- read_csv("./2016_US_County_Level_Presidential_Results.csv")


election16 <- electionRaw %>% 
  mutate(
    county_fips = combined_fips
  ) %>% 
  select(county_fips, 2:7) %>% 
  mutate(
    trump = case_when(per_gop > per_dem ~ 1,
                      per_dem > per_gop ~ 0,
                      per_gop == per_dem ~ -999,
                      .default = NA),
    bell = if_else(trump == 1, 1, 0)
  )

table(election16$trump, useNA = c("always"))

rm(electionRaw)
str(election16)
```

## Clean and Combine

```{r}

# Deal with special case for Connecticut which changed from counties to Planning regions in 2024




election <- election24 %>% 
  left_join(election20, by = join_by(county_fips), suffix = c("_24", "")) %>%
  mutate(
    GEOID = county_fips,
    county_fips = as.numeric(county_fips)
  ) %>% 
  left_join(election16, by = join_by(county_fips), suffix = c("_20", "_16")) %>% 
  mutate(
    bellweather = ifelse(bell_16 == 1 & bell_20 == 1 & bell_24 == 1, 1, 0),
    pop_vote_bell = ifelse(trump_16 == 0 & trump_20 == 0 & trump_24 == 1, 0, 1),
    bell_20_24 = ifelse(bell_20 == 1 & bell_24 == 1, 1, 0),
    change = ifelse(trump_16 == trump_20 & trump_20 == trump_24, 0, 1),
    change_20_24 = ifelse(trump_20 == trump_24, 0, 1)
  ) %>% 
  filter(!(state_name %in% c("Alaska","Connecticut", "District of Columbia")))

table(election$change)
table(election$bellweather)
table(election$pop_vote_bell)

#rm(election16, election20, election24)

```
```{r}
write_csv(election, "./cleanData/election.csv")
```





## Covid

There are 12 Datasets available with the COVID data

1. Consumer Spending

```{r}
spendingRaw <- read_csv("./EconomicTracker-main/data/Affinity - County - Daily.csv")

#str(spendingCounty)

#spendingNational <- read_csv("./EconomicTracker-main/data/Affinity - National - Daily.csv")

#str(spendingNational)

spendingRaw %>% 
  #filter(year == 2022) %>% 
  mutate(
    spend_all = ifelse(spend_all == ".", NA, spend_all),
    missing = ifelse(is.na(spend_all), 1, 0)
  ) %>% 
  group_by(countyfips) %>% 
  filter(countyfips==1001)

```
```{r}

pad_zero <- function(x) {
  x %>%
    as.character() %>%
    if_else(nchar(.) == 1, paste0("0", .), .)
}

spend <- spendingRaw %>% 
  mutate(
    spend_all = ifelse(spend_all == ".", NA, spend_all),
    missing = ifelse(is.na(spend_all), 1, 0)
  ) %>% 
  mutate(
    month = pad_zero(month),
    day = pad_zero(day)
  ) %>% 
  unite("date", year, month, day, sep = "", remove = FALSE) %>% 
  mutate(
    date = ymd(date),
    woy = week(date),
    dow = wday((date))
  ) %>% 
  filter(dow==1)%>% 
  mutate(
    spend_all = as.numeric(spend_all)
  ) 
  

  
```

```{r}
spend %>% 
  filter(!is.na(spend_all))# %>% 
  # group_by(countyfips) %>% 
  # mutate(
  #   nobs = n()
  # ) %>% 
 # count(countyfips)#%>% 
  #distinct(n)
  #()
```

```{r}
spendingRaw %>% 
  filter(countyfips %in% c(48003, 48227, 48475, 39051)) %>% 
  arrange(countyfips) %>% 
  view()
```


```{r}
summarise_changes <- function(data, column, tag) {
  column_sym <- ensym(column)         # capture column name as symbol
  column_name <- tag # get the string name
  
  data %>%
    summarise(
      periods = n(),
      !!paste0("final_", column_name) := last(!!column_sym),
      !!paste0("average_", column_name) := mean(!!column_sym, na.rm = TRUE),
      !!paste0("sd_", column_name) := sd(!!column_sym, na.rm = TRUE),
      !!paste0("positive_", column_name) := sum(!!column_sym > 0, na.rm = TRUE),
      !!paste0("negative_", column_name) := sum(!!column_sym < 0, na.rm = TRUE),
      !!paste0("max_increase_", column_name) := max(!!column_sym, na.rm = TRUE),
      !!paste0("max_decrease_", column_name) := min(!!column_sym, na.rm = TRUE),
      .groups = "drop"
    )
}
```




```{r}
spend %>% 
  mutate(
    spend_all = as.numeric(spend_all)
  ) %>% 
  filter(!is.na(spend_all)) %>% 
  filter(year==2020) %>% 
  group_by(woy) %>%
  summarise(
    count = n(),                  # number of observations
    mean_spend = mean(spend_all),  # mean of spend_all
    min_spend = min(spend_all),
    quan25 = quantile(spend_all, 0.25),
    median = median(spend_all),
    quan75 = quantile(spend_all, 0.75),
    max_spend = max(spend_all), 
    date = first(date), # maximum of spend_all
    .groups = "drop"               # ungroup after summarizing
)
```
```{r}
# Summarize by countyfips
spendingFinal <- spend %>%
  filter(!is.na(spend_all)) %>% 
  group_by(countyfips) %>%
  mutate(
    check = n()
  ) %>% 
  filter(check > 200) %>% 
  # filter(countyfips==1001) %>% 
  #summarise(mean = mean(spend_all))
  # summarise(
  #   periods = n(),
  #   final_change = last(spend_all),
  #   average_change = mean(spend_all, na.rm = TRUE),
  #   sd_change = sd(spend_all, na.rm = TRUE),
  #   positive_periods = sum(spend_all > 0),
  #   negative_periods = sum(spend_all < 0),
  #   max_increase = max(spend_all, na.rm = TRUE),
  #   max_decrease = min(spend_all, na.rm = TRUE),
  #   .groups = "drop"  # Ungroup after summarizing
  # )
  summarise_changes(spend_all, "consumer_spending")
# Print summary
head(spendingFinal)

write_csv(spendingFinal,"./cleanData/consumerSpending.csv")

```
2. Small Business
```{r}
businessRaw <- read_csv("./EconomicTracker-main/data/Womply - County - Weekly.csv")
```

```{r}
sbOpen <- businessRaw %>% 
  group_by(countyfips) %>% 
  summarise_changes(merchants_all, "sb_open")

sbRevenue <- businessRaw %>% 
  group_by(countyfips) %>% 
  summarise_changes(revenue_all, "sb_rev")
```

```{r}
businessRaw %>% 
  select(countyfips) %>% 
  distinct()
```



## Job Postings
```{r}
jobRaw <- read_csv("./EconomicTracker-main/data/Job Postings - State - Weekly.csv")
```
```{r}
job <- jobRaw %>% 
  filter(year < 2025) %>% 
  filter(!(year == 2024 & month >= 11)) %>% 
  mutate(
    bg_posts = ifelse(bg_posts == ".", NA, bg_posts)
  )

job %>% filter(!is.na(bg_posts))
```

Too much missing data, not gonna use this.


## Employment

```{r}
rm(job, jobRaw)
employmentRaw <- read_csv("./EconomicTracker-main/data/Employment - County - Weekly.csv")
```
```{r}

  
employmentRaw[employmentRaw == "."] <- NA


employment <- employmentRaw %>% 
  filter(!is.na(emp)) %>% 
  filter(year < 2025) %>% 
  filter(!(year == 2024 & month >= 11)) %>% 
  mutate(
    emp = as.numeric(emp)
  )


employmentClean <- employment %>% 
  group_by(countyfips) %>% 
  summarise_changes(emp, "employment")

```

## Unemployment

```{r}

uiCountyRaw <- read_csv("./EconomicTracker-main/data/UI Claims - County - Weekly.csv")

uiCountyRaw[uiCountyRaw == "."] <- NA 

uiCounty <- uiCountyRaw %>% 
  filter(!is.na(initclaims_rate_regular))%>% 
  filter(year < 2025) %>% 
  filter(!(year == 2024 & month >= 11)) %>% 
  mutate(
    initclaims_rate_regular = as.numeric(initclaims_rate_regular)
  )
  
uiCountyClean <- uiCounty %>% 
  group_by(countyfips) %>% 
  summarise_changes(initclaims_rate_regular, "ui_claims_rate") %>% 
  select(-6:-7)




```

```{r}
covidRaw1 <- read_csv("./EconomicTracker-main/data/COVID - County - Daily 2020.csv")

covidRaw2 <- read_csv("./EconomicTracker-main/data/COVID - County - Daily 2021.csv")

covidRaw3 <- read_csv("./EconomicTracker-main/data/COVID - County - Daily 2022.csv")

covidRaw <- covidRaw1 %>% 
  bind_rows(covidRaw2, covidRaw3)


covidRaw[covidRaw == "."] <- NA 


covid <- covidRaw %>% 
  mutate(
    month = pad_zero(month),
    day = pad_zero(day)
  ) %>% 
  unite("date", year, month, day, sep = "", remove = FALSE) %>% 
  mutate(
    date = ymd(date),
    woy = week(date),
    dow = wday((date))
  ) %>% 
  filter(dow==1) %>% 
  mutate(across(5:29, as.numeric))


covid  %>% 
  filter(countyfips==6057)
  #filter(!is.na(case_count))


covid %>%
  
  filter(countyfips==6057) %>% str()
  #filter(case_count>0)


covidClean <- covid %>% 
  group_by(countyfips) %>% 
  summarize(
    final_death_rate = max(death_rate, na.rm=TRUE),
    final_case_rate = max(case_rate, na.rm=TRUE),
    max_new_death_rate = max(new_death_rate, na.rm = TRUE),
    max_new_case_rate = max(new_case_rate, na.rm = TRUE),
    weeks = sum(!is.na(case_count)),
    avg_new_case_rate = sum(new_case_rate, na.rm = TRUE) / weeks,
    avg_new_death_rate = sum(new_death_rate, na.rm = TRUE) / weeks,
  ) %>% 
  mutate(across(2:8, ~ifelse(weeks==0, 0, .x)))

```
Google Mobility


```{r}
googleRaw <- read_csv("./EconomicTracker-main/data/Google Mobility - County - Daily.csv")


google <- googleRaw %>% 
  mutate(
    month = pad_zero(month),
    day = pad_zero(day)
  ) %>% 
  unite("date", year, month, day, sep = "", remove = FALSE) %>% 
  mutate(
    date = ymd(date),
    woy = week(date),
    dow = wday((date))
  ) %>% 
  filter(date >= ymd("2020-03-01") & date <= ymd("2022-12-31"))


google %>% 
  filter(countyfips==1023)
#google[google == "."] <- NA 

googleClean <- google %>% 
  group_by(countyfips) %>% 
  mutate(
    google = as.numeric(gps_away_from_home)
  ) %>% 
  filter(!is.na(google)) %>% 
  summarise_changes(google, "away_from_home") %>% 
  mutate(
    percent_positive = positive_away_from_home/ (positive_away_from_home + negative_away_from_home),
    .keep = "unused"
  )

```

# Join all datasets

```{r}
countyData <- election %>% 
  left_join(covidClean, by =join_by(county_fips==countyfips)) %>% 
  left_join(employmentClean, by =join_by(county_fips==countyfips)) %>%
  left_join(googleClean, by =join_by(county_fips==countyfips)) %>% 
  left_join(sbOpen, by =join_by(county_fips==countyfips)) %>% 
  left_join(sbRevenue, by =join_by(county_fips==countyfips)) %>% 
  left_join(spendingFinal, by =join_by(county_fips==countyfips)) %>% 
  left_join(uiCountyClean, by =join_by(county_fips==countyfips)) %>% 
  select(-starts_with("periods"))
```




## Imputing Missing Data

just use state, and fill in for missing counties, use the same variables as available for county data. This 
makes it an apples to apples comparison. For swing states / bellwether counties, explore this in EDA, just
use classification models for predicting 2024 across country. 






## Demographic

Get 2010 and 2020, as well as ACS
- Age
- Race
- Sex
- Education
- Income


