---
title: "Modeling"
author: "David Hyon"
date: "2025-04-28"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}

set.seed(123)


# Load necessary library
library(e1071)
library(tidyverse)

election <- read_csv("./modelData.csv") %>% 
  select(state_name, county_name, everything()) %>% 
  select(
    -GEOID,
    -state_name_24,
    -county_name_24,
    -statefips
  )



# Make sure the response is a factor for classification
election$trump_24 <- as.factor(election$trump_24)

# Create the predictor matrix and response vector
X <- election[, 33:169]
y <- election$trump_24


dat <- data.frame(x = X, y = y)

train <- sample(3100, 3100*0.8)

trainDat <- dat[train,]
testDat <- dat[-train,]

```


# SVM

```{r}

# Build the SVM model
svm_model_linear <- svm(y ~ ., data = trainDat, 
                 kernel = "linear",   # Try "radial" if nonlinear
                 scale = TRUE)        # Standardize predictors

# View model summary
summary(svm_model_linear)
```



```{r}

set.seed(123)
#plot(svm_model_linear, dat[,1:2])


# tune.out <- tune(svm, y∼., data = dat, kernel = "linear",
#   ranges = list(cost = c(0.001, 0.01, 0.1, 1, 5, 10, 100)))

# tune.out <- tune(svm, y ~ ., data = trainDat, kernel = "linear", probability = TRUE, 
#                  ranges = list(cost = c(0.001, 0.01, 0.1, 1, 5)))
```

```{r}
# summary(tune.out)
```
```{r}
set.seed(123)

svc <- svm(y ~ ., data = trainDat, 
                 kernel = "linear",
            probability = TRUE,
           cost = 0.01,
           # Try "radial" if nonlinear
                 scale = TRUE)


ypred <- predict(svc, testDat)
table(predict = ypred, truth = testDat$y)

table(dat$y)

```
Misclassification rate of 7%. 93% accuracy. There is a data imbalance, so 85% of counties voted for trump. Still, this is better than random. 

Time for a support vector machine

```{r}
# set.seed(123)
# tune.out_radial <- tune(svm, y ~ ., data = trainDat, kernel = "radial",  probability = TRUE,
#                         ranges = list(cost = c(0.1, 1,10, 100),
#                                       gamma = c(0.5, 1, 2, 3, 4)))
```

```{r}
# summary(tune.out_radial)
```
```{r}
# summary(tune.out_radial$best.model)


svm_radial <- svm(y ~ ., data = trainDat, 
                 kernel = "radial",
            probability = TRUE,
           cost = 0.01,
           # Try "radial" if nonlinear
                 gamma = 0.5)

ypred_radial <- predict(svm_radial, testDat)
table(predict = ypred_radial, truth = testDat$y)
```
So the SVC does a better job than SVM with radial kernel. 


# Tree-based methods
```{r}
set.seed(123)

library(tree)

tree.data <- tree(y ~ .,dat,subset = train)
tree.pred <- predict(tree.data, testDat, type = "class")
table(tree.pred, y[-train])


```
10% Error. Better than random. But having a tough time predicting democrat counties. 


```{r}

set.seed(123)

cv.data <- cv.tree(tree.data, FUN = prune.misclass)
names(cv.data)
cv.data
```
```{r}
par(mfrow = c(1, 2))
plot(cv.data$size, cv.data$dev, type = "b")
plot(cv.data$k, cv.data$dev, type = "b")
```

```{r}
set.seed(123)


prune.data <- prune.misclass(tree.data, best = 8)
plot(prune.data)
text(prune.data, pretty = 0)
```
```{r}
set.seed(123)

tree.pred <- predict(prune.data, testDat, type = "class")
table(tree.pred, y[-train])
```
Basically the same error, but with less nodes. 

## Random Forest

```{r}

set.seed(123)

library(randomForest)

rf.data <- randomForest(y ~ ., data = trainDat, importance = TRUE)
rf.pred <- predict(rf.data, newdata = testDat, type = "class")
table(rf.pred, y[-train])

```
7%, on par with linear SVM

```{r}
#importance(rf.data)
```

```{r}
varImpPlot(rf.data, n.var = 10)
```
```{r}
# Extract importance
var_imp <- importance(rf.data)
var_imp_df <- as.data.frame(var_imp)
var_imp_df$Variable <- rownames(var_imp_df)

# Sort and select top 8
accuracy <- var_imp_df[order(var_imp_df$MeanDecreaseAccuracy, decreasing = TRUE), ][1:10, ]
gini <- var_imp_df[order(var_imp_df$MeanDecreaseGini, decreasing = TRUE), ][1:10, ]

varNameAcc <- c("% White", "% Black", "Masters Degree %", "Professional Degree %", "Max COVID-19 Weekly Death Rate", "Population", "PhD %", "% Asian", "COVID-19 Death Rate", "High School Diploma %")


varNameGini <- c("% White", "Masters Degree %","% Black",   "Professional Degree %",  "Max COVID-19 Weekly Death Rate","PhD %","% Asian", "High School Diploma %", "Population", "Bachelors Degree %")

# Plot with ggplot2
accPlot <- ggplot(accuracy, aes(x = reorder(varNameAcc, MeanDecreaseAccuracy), y = MeanDecreaseAccuracy)) +
  geom_point(col = "steelblue") +
  coord_flip() +  # Horizontal bars
  labs(#title = "Top 8 Variable Importances",
       x = "",
       y = "Mean Decrease Accuracy") #+
  #theme_minimal()

giniPlot <- ggplot(gini, aes(x = reorder(varNameGini, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_point(col = "steelblue") +
  coord_flip() +  # Horizontal bars
  labs(#title = "Top 8 Variable Importances",
       x = "",
       y = "Mean Decrease GINI")


gridExtra::grid.arrange(accPlot, giniPlot, ncol=2)
```
## Boosted Trees

```{r}
set.seed(123)

library(gbm)

trainDatBoost <- trainDat %>% 
  mutate(
    y = as.numeric(y)-1
  )

testDatBoost <- testDat %>% 
  mutate(
    y = as.numeric(y)-1
  )


boost.data <- gbm(y ~ ., data = trainDatBoost, distribution = "bernoulli", 
                  n.trees = 5000, interaction.depth = 4)
                  
summary(boost.data)

```
```{r}
# 1. Get relative influence WITHOUT plotting
rel_influence <- summary(boost.data, plotit = FALSE)

# 2. Take top 10 variables
top10 <- rel_influence[order(-rel_influence$rel.inf), ][1:10, ]

boostVarNames <- c("% White", "Masters Degree %", "Professional Degree %", 
                   "Max COVID-19 Weekly Death Rate", "COVID-19 Death Rate","% Black","Population", "% Asian",    "Some College Less than 1 year", "% American Indian/Alaska Native")

 

# 3. Plot with ggplot2
ggplot(top10, aes(x = reorder(boostVarNames, rel.inf), y = rel.inf)) +
  geom_col(fill = "steelblue") +
  coord_flip() +  # Horizontal bars
  labs(
    #title = "Top 10 Variable Influences",
    x = " ",
    y = "Relative Influence (%)"
  ) +
  theme_minimal(base_size = 14)
```
```{r}
plot(boost.data, i = "x.one_race_white_alone")
plot(boost.data, i = "x.estimate_total_masters_degree")
```
```{r}
set.seed(123)


yhat.boost <- predict(boost.data, newdata = testDatBoost, n.trees = 5000, type = "response")
rf.pred <- predict(rf.data, newdata = testDat, type = "class")

boost.pred <- ifelse(yhat.boost > 0.5, 1, 0)

table(boost.pred, y[-train])

show.gbm(boost.data)
```

5% misclassification rate. Best so far. 

## Logistic Regression with Lasso

```{r}
set.seed(123)
library(glmnet)

X_lasso <- as.matrix(trainDat[, colnames(trainDat) != "y"])
y_lasso <- trainDat$y

# Step 2: Fit Lasso logistic regression with cross-validation
lasso_fit <- cv.glmnet(
  X_lasso, y_lasso, 
  family = "binomial",
  alpha = 1,       # Lasso penalty
  nfolds = 5
)

# Step 3: Find best lambda
best_lambda <- lasso_fit$lambda.min

# Step 4: Extract selected variables
lasso_coefs <- coef(lasso_fit, s = best_lambda)
selected_vars <- rownames(lasso_coefs)[lasso_coefs[,1] != 0]
selected_vars <- setdiff(selected_vars, "(Intercept)")

# Step 5: Build logistic regression on selected variables
# Subset the original data to selected variables
X_selected <- trainDat[, selected_vars, drop = FALSE]

# Build final logistic regression
final_logit_model <- glm(
  y ~ ., 
  data = data.frame(y = y_lasso, X_selected), 
  family = "binomial"
)

# Step 6: View model summary
summary(final_logit_model)
```
```{r}
set.seed(123)

# 1. Prepare test data
X_test_selected <- testDat[, selected_vars, drop = FALSE]

# 2. Predict probabilities
prob_lasso <- predict(final_logit_model, newdata = X_test_selected, type = "response")

# 3. Predict classes
pred_lasso <- ifelse(prob_lasso > 0.5, 1, 0)
```










# Model Comparison

```{r}

set.seed(123)

# Load needed libraries
library(caret)
library(pROC)

# # Define function to calculate evaluation metrics
# evaluate_model <- function(pred_class, actual_class, pred_probs) {
#   conf_matrix <- confusionMatrix(pred_class, actual_class, positive = "1")
#   
#   precision <- conf_matrix$byClass["Pos Pred Value"]
#   recall <- conf_matrix$byClass["Sensitivity"]
#   f1 <- 2 * (precision * recall) / (precision + recall)
#   
#   roc_obj <- roc(actual_class, pred_probs)
#   auc_value <- auc(roc_obj)
#   
#   metrics <- list(
#     Accuracy = conf_matrix$overall["Accuracy"],
#     Precision = precision,
#     Recall = recall,
#     F1_Score = f1,
#     AUC = auc_value
#   )
#   
#   return(metrics)
# }




evaluate_model <- function(pred_class, actual_class, pred_probs) {
  conf_matrix <- confusionMatrix(pred_class, actual_class, positive = "1")
  
  precision <- as.numeric(conf_matrix$byClass["Pos Pred Value"])
  recall    <- as.numeric(conf_matrix$byClass["Sensitivity"])
  f1        <- 2 * (precision * recall) / (precision + recall)
  
  roc_obj   <- roc(actual_class, pred_probs)
  auc_value <- as.numeric(auc(roc_obj))
  
  metrics <- list(
    Accuracy  = as.numeric(conf_matrix$overall["Accuracy"]),
    Precision = precision,
    Recall    = recall,
    F1_Score  = f1,
    AUC       = auc_value
  )
  
  return(metrics)
}


```




```{r}
set.seed(123)


# Actual Test values as Factor
yTest <- y[-train]


## Boost Model

# # Probabilities
prob_boost <- yhat.boost

# Predicted Class
pred_boost <- as.factor(boost.pred)


eval_boost <- evaluate_model(pred_boost, yTest, prob_boost)


## RF Model

# Probabilities
prob_rf <- predict(rf.data, newdata = testDat, type = "prob")[,1]

# Predicted Class
pred_rf <- rf.pred

eval_rf <- evaluate_model(pred_rf, yTest, prob_rf)



## CV Single Tree

# Probabilities
prob_tree <- predict(tree.data, newdata = testDat, type = "vector")[,1]

# Predicted Class
pred_tree <- tree.pred

eval_tree <- evaluate_model(pred_tree, yTest, prob_tree)



## Refit Logistic Regression

# Predicted Class
pred_lasso <- as.factor(pred_lasso)

eval_lasso <- evaluate_model(pred_lasso, yTest, prob_lasso)


## SVC
# Predict on test data
pred_svc <- predict(svc, newdata = testDat, probability = TRUE)

# Extract predicted probabilities
prob_svc <- attr(pred_svc, "probabilities")[,1]

# Predicted class
eval_svc <- evaluate_model(pred_svc, yTest, prob_svc)


## SVM
# Predict on test data
pred_svm <- predict(svm_radial, newdata = testDat, probability = TRUE)

# Extract predicted probabilities
prob_svm <- attr(pred_svm, "probabilities")[,1]

# Predicted class
eval_svm <- evaluate_model(pred_svm, yTest, prob_svm)

```


```{r}


model_evals <- rbind(eval_boost, eval_rf, eval_tree, eval_lasso, eval_svc, eval_svm) %>%
  cbind(Model = c("Boosted Tree", "Random Forest", "Decision Tree", "Log Regression", "SVM (Linear)", "SVM (Radial)")) %>%
as_tibble() %>%
select(Model, everything()) %>%
  mutate(across(where(is.list), ~ unlist(.)))

write_csv(model_evals, "./cleanData/evals.csv")
  
```


```{r}
knitr::kable(model_evals, caption = "Model Evaluation Metrics")
```



## Plot ROC Curves



```{r}
library(pROC)
library(ggplot2)



roc_boost <- roc(yTest, prob_boost, levels = c("1", "0"))
roc_rf    <- roc(yTest, prob_rf, levels = c("1", "0"))
roc_tree  <- roc(yTest, prob_tree, levels = c("1", "0"))
roc_lasso <- roc(yTest, prob_lasso, levels = c("1", "0"))
roc_svc   <- roc(yTest, prob_svc, levels = c("1", "0"))
roc_svm   <- roc(yTest, prob_svm, levels = c("1", "0"))

# Combine and label with AUC
roc_list <- list(
  "Boosted Tree"  = roc_boost,
  "Random Forest" = roc_rf,
  "Decision Tree" = roc_tree,
  "Logistic Regression" = roc_lasso,
  "SVC" = roc_svc,
  "SVM" = roc_svm
)

# Rename with AUC values included
roc_list_named <- setNames(
  roc_list,
  sapply(names(roc_list), function(name) {
    auc_value <- auc(roc_list[[name]])
    sprintf("%s (AUC = %.2f)", name, auc_value)
  })
)




# Create the ROC plot
ggroc(roc_list_named, aes = c("color"), linewidth = 1.2, legacy.axes = TRUE) +
  scale_color_brewer(palette = "Set1") +
  labs(
    x = "False Positive Rate",
    y = "True Positive Rate",
    title = "ROC Curve Comparison",
    color = "Model"
  ) +
  scale_color_brewer(palette = "Set1", direction = -1) +
  geom_abline(linetype = "dashed", color = "grey50") 
```

